An Algorithm for Searching Evolving Graphs in Linear Algebraic Language

### 2016-01-23 paper.tex
Simplify description of implementation.
Take out future work from conclusion.

Better acks.

To study evolving graphs and experiment with various graph types, we have developed EvolvingGraphs.jl  \cite{zhang15}, a software package for the creation, manipulation, and study of evolving graphs written in Julia \cite{bkse12}. The implementation of the evolving-graph BFS is
available online at
\begin{center}
\url{https://github.com/weijianzhang/EvolvingGraphs.jl}.
\end{center}
%Julia is a high-level, high-performance dynamic programming language for technical
%computing.
% We make particular use of
% the parameterizable type system and multiple dispatch in Julia.
 \texttt{MatrixList}, a data type in EvolvingGraphs.jl,  represents an evolving graph
as a sequence of sparse matrices stored in compressed sparse column (CSC) format.
Many basic graph functions are implemented for \texttt{MatrixList}.

@@

We shall map a database of research papers to an evolving graph $G_n$
in such a way that if author $i$ cited author $j$ on a paper published
at  time $t$, the evolving graph $G_n$ contains a directed edge from
$i$ to $j$ at time stamp $t$. Then given an author $a$ at time stamp $t_0$,
we can use our evolving-graph BFS to construct $T(a, t_0)$ which is
a set of all the authors that have been influenced by $a$'s work at
time $t_0$. We define a \emph{community} to be a group of
researchers that have been influenced by the same authors.
For example, given a paper published by $a$ at year $2016$,
we can determine $a$'s community by first backward searching to
find the root authors then forward searching to find the descendants.

To design such information retrieval system, two parameters will be needed:
 one for determining the layers of the backward search tree which can be
adjusted by the user;
the other for determining root authors for forward searching
so that we can
filter out unessential or overlapped trees, which can be computed
by ranking algorithms like PageRank \cite{pbmw99} or machine-learning based preference ranking.

Tracing influence through citation networks naturally lead to considering paths
through evolving graphs that traverse time backwards. While we have until now
only considered forward evolution in time, the treatment for reverse evolution
follows naturally simply by reversing the time labels, e.g.\ by the transformation
$t\rightarrrow -t$.

@@

In this paper, we have introduced and analyzed the evolving-graph BFS in both
graph and algebraic notation. The relationship between active nodes and
evolving graph algorithms is discussed.
This paper leaves important extensions open for further research. First, it is
of interest to extend the evolving-graph BFS algorithm for weighted evolving graphs, which can provide more valuable information when the degree of connectivity is available in the data.
Second, efficient algebraic representations and algorithms need to be addressed.
One major advantage of representing evolving graphs as sparse matrices
is the matrix representation is highly structured so it is easy to design
high performance parallel implementation.
Third, depth-first search and other graph traversal algorithms can be
extended using similar ideas.

\section*{Acknowledgment}

The authors are grateful to Professor Nicholas J. Higham and Doctor Vedran \u Sego for suggestions.
This work was completed while the second author visited MIT CSAIL. He thanks
Professor Alan Edelman for making his stay so pleasant.

>>>>

To study evolving graphs and experiment with various graph types, we have
developed EvolvingGraphs.jl~\cite{zhang15}, a software package for the creation,
manipulation, and study of evolving graphs written in Julia~\cite{bkse12}.
The implementation of the evolving graph BFS is
freely available online at
\begin{center}
\url{https://github.com/weijianzhang/EvolvingGraphs.jl}
\end{center}
and available with the MIT "Expat" license.

%Julia is a high-level, high-performance dynamic programming language for technical
%computing.
% We make particular use of
% the parameterizable type system and multiple dispatch in Julia.
\texttt{MatrixList}, a data type in EvolvingGraphs.jl, represents an evolving graph
as a sequence of sparse matrices stored in compressed sparse column (CSC) format.
Many basic graph functions are also implemented for \texttt{MatrixList}.

@@

In this section, we show that evolving graph formalism presented above can be
used to capture the structure of citation networks. Consider
the evolving graph $G_n=\langle G^{[t]} \rangle_t$ such that $G^{[t]}$ has
node set corresponding to authors active at time $t$ and directed edge set
$E^{[t]} \ni (i, j)$ representing a citation of author $j$ by author $i$ in a
publication at time $t$.

Then given an author $a$ at time $t_1$, the evolving graph BFS described above
can compute $T(a, t_1)$, the set of all the authors that have been influenced by
$a$'s work at time $t_1$.
Define also a \emph{community} to be a group of
researchers that have been influenced by the same authors.
For example, given a paper published by $a$ at time $t$,
we can determine $a$'s community by searching backward in time to find
$T^-1(a, t_1)$, the authors that influenced $a$ at time $t$, and also searching
forward to find $T(a, t_1)$.
The backward search in time follows straightforwardly from the forward time traveral
presented above simply by reversing the time labels, e.g.\ by the transformation
$t\rightarrow -t$.

@@

This paper introduced the generalization of BFS to evolving graphs and presented
an equivalent formulation using linear algebra. The concept temporal paths
connecting active nodes turns out to be critical for the correct generalization
of BFS to evolving graphs. An efficient implementation of BFS on evolving
graphs is now available as a open source Julia package.

% This paper leaves important extensions open for further research. First, extensions
% of BFS to weighted evolving graphs can provide more valuable information when the degree of connectivity is available in the data.
% Second, efficient algebraic representations and algorithms need to be addressed.
% One major advantage of representing evolving graphs as sparse matrices
% is the matrix representation is highly structured so it is easy to design
% high performance parallel implementation.
% Third, depth-first search and other graph traversal algorithms can be
% extended using similar ideas.

\section*{Acknowledgments}

We thank N. J. Higham and V. \u Sego (Manchester) for helpful suggestions.
W.\ Z.\ thanks the School of Mathematics at U. Manchester for travel funding and
A. Edelman for arranging for a fruitful visit to MIT.

%XXX ANY OTHER FUNDING
